---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# aurum

<!-- badges: start -->
<!-- badges: end -->

The goal of aurum is to ...

## Installation

You can install the released version of aurum from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("aurum")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("drkgyoung/CPRD_Aurum")
```
## Example


## Configruation



### Querying

`~/.aurum.yaml`

```YAML
default:
  user: <mysql username>
  password: <sql password>
  server: "localhost"
  port: 3306

analysis:
  dataDatabase: cprd_data
  analysisDatabase: cprd_analysis

test-analysis:
  dataDatabase: cprd_data
  analysisDatabase: cprd_analysis_dev

test-remote:
  dataDatabase: cprd_data
  analysisDatabase: cprd_analysis_dev
  server: 127.0.0.1
  port: 3307
```

### Database loading configuration

During the database load phase some additional configuration was required to setup the data directories
This is `~/.aurum.load.yaml`

```YAML
default:
  user: <mysql username>
  password: <sql password>
  server: "localhost"
  port: 3306
  lookupSourceDirectory: "/slade/DBs/mysql-files/CPRD/reference-data/202005_Lookups_CPRDAurum"
  dataSourceDirectories:
    - /slade/DBs/mysql-files/CPRD/master_males
    - /slade/DBs/mysql-files/CPRD/master_females

dev:
  dataDatabase: cprd_data_dev
  analysisDatabase: cprd_analysis_dev
  buildDirectory: /slade/DBs/mysql-files/CPRD/staging-dev
  loadLimit: 10

prod:
  dataDatabase: cprd_data
  analysisDatabase: cprd_analysis
  buildDirectory: /slade/DBs/mysql-files/CPRD/staging
  loadLimit: 100000

test-analysis:
  dataDatabase: cprd_data
  analysisDatabase: cprd_analysis_dev
  buildDirectory: /slade/DBs/mysql-files/CPRD/staging
  loadLimit: 10


```

## Connection and querying

This is a basic example which shows you how to solve a common problem:

```{r example}
library(aurum)
## basic example code
```


